Model Recognition as Planning

### 2018-10-12 doc/icaps-19/model-recognition.tex
new full version icaps19 ready for review

\maketitle
\begin{abstract} 
Given the partial observation of a plan execution and a set of possible planning models (models that share the same state variables but that update these variables according to different action models), {\em model recognition} is the task of identifying which model in the set explains/produced the given observation. The paper formalizes the {\em model recognition} task and introduces a novel method to estimate the probability of a \strips\ model to produce an observation of a plan execution. This method, that we called {\em model recognition as planning}, is built on top of off-the-shelf classical planning algorithms and is robust to missing intermediate states and actions in the observed plan execution. The effectiveness of {\em model recognition as planning} is shown in a set of \strips\ models encoding different kinds of {\em automata}. We show that {\em model recognition as planning} succeeds to identify the executed automata despite the internal machine state or actual applied transitions, are unobserved.
\end{abstract}

@@

This paper formalizes the {\em model recognition} task where the object to recognize is not a goal but the {\em planning model} that determines the behavior of the observed agent. Given a partial observation of a plan execution and a set of possible planning models (models that share the same state variables but that update these variables with different action models), {\em model recognition} is the task of identifying the model in the set with the highest probability of producing/explaining the given observation.

\begin{figure}
  \begin{scriptsize}    
  \begin{center}
  \begin{tikzpicture}[scale=.4]
          \begin{scope}
            \draw (0, 0) grid (5, 5);
            \node[anchor=center] at (3.5, 1.5) {R};
            \node[anchor=center] at (0.5, 0.5) {};
            \draw[thick,style=dotted] (0.5,0.5) -- (4.5,0.5);
            \draw[thick,style=dotted] (4.5,0.5) -- (4.5,1.5);
            \draw[thick,style=dotted,->] (4.5,1.5) -- (3.7,1.5);                        
          \end{scope}
        \end{tikzpicture}
\hspace*{1cm}        
	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,semithick]
	  \node[state] (A)              {$q_0$};
	  \node[state] (B) [right of=A] {$q_1$};          
	  \path
(A) edge [bend left, align=center] node {inc-y,\\dec-y} (B)
(B) edge [bend left, align=center] node {inc-y,\\dec-y} (A);
	\end{tikzpicture}
  \end{center}
  \end{scriptsize}  
 \caption{\small ({\em Left)} Robot navigating a $5\times 5$ grid. {\em (Right)} Automata controling that the robot only increments its x-coordinate at {\em even} rows (when $q_0$ holds, given that actions {\tt inc-y} and {\tt dec-y} update the robot y-coordinate and the automata state).}
\label{fig:grid-example}
\end{figure}

To better illustrate {\em model recognition}, imagine a robot in a $n\times n$ grid whose navigation is determined by the \strips\ model of Figure~\ref{fig:model-example}. According to this model the robot could increment its {\em x-coordinate} at {\em even rows} (when {\tt\small $q0$} holds) and decrement it at the {\em odd rows} (when {\tt\small $q1$} holds). Apart from this particular navigation model, different action models could be defined within the same state variables (e.g. altering the way {\tt\small $q0$} and {\tt\small $q1$} are required and updated) and these models can determine different kinds of robot navigation. Given an observation of a plan execution (like the one illustrated at Figure~\ref{fig:grid-example}) {\em model recognition} would aim here to identify which navigation model produced/explains that observation, despite key information is unobserved (e.g. the particular applied actions or the value of the state variables {\small\tt $q0$} and {\small\tt $q1$}).

{\em Model recognition} is of interest because once the planning model is recognized, then the model-based machinery for automated planning becomes applicable~\cite{ghallab2004automated}. In addition, it enables identifying different kinds of automata by observing their execution. It is well-known that diverse automatae representations, like {\em finite state controllers}, {\em push-down automata}, {\em {\sc GOLOG} programs} or {\em reactive policies}, can be encoded as classical planning models~\cite{baier2007exploiting,Geffner:FSM:AAAI10,ivankovic2015optimal,segovia2017generating}.

The paper also introduces {\em model recognition as planning}; a novel method to estimate the probability of a given \strips\ model to produce an observed plan execution. Our method is built on top of off-the-shelf classical planning algorithms and is robust to missing intermediate states and actions in the observed plan execution.

We evaluate the effectiveness of {\em model recognition as planning} with sets of \strips\ models that represent different {\em automata}. All the {\em automata} in a set are defined within the same state variables (same input {\em alphabet} and same {\em machine states}) but different {\em transition functions}. We show that {\em model recognition as planning} succeeds to identify the executed {\em automata} despite internal machine states or actual applied transitions are unobserved.

@@

A {\em classical planning frame} is a $\tup{F,A}$ pair, where $F$ is a set of {\em fluents} and $A$ is a set of {\em actions}. The semantics of actions $a\in A$ are specified with two functions: $\rho(s,a)$ that denotes whether the action is {\em applicable} in a state $s$ and $\theta(s,a)$ that denotes the {\em successor state} that results of applying the action $a$ in a state $s$. In this work we specify action semantics (the $\rho$ and $\theta$ functions) with the \strips\ model. With this regard, an action $a\in A$ is defined by:
\begin{itemize}
\item $\pre(a)\in\mathcal{L}(F)$, the {\em preconditions} of $a$, is the set of literals that must hold for the action $a\in A$ to be applicable.
\item $\eff^+(a)\in\mathcal{L}(F)$, the {\em positive effects} of $a$, is the set of literals that are true after the application of action $a\in A$.
\item $\eff^-(a)\in\mathcal{L}(F)$, the {\em negative effects} of $a$, is the set of literals that are false after the application of the action.
\end{itemize}
We say that an action $a\in A$ is {\em applicable} in a state $s$ iff $\pre(a)\subseteq s$. The result of applying $a$ in $s$ is the {\em successor state} denoted by $\theta(s,a)=\{s\setminus\eff^-(a))\cup\eff^+(a)\}$.

@@

\subsection{Conditional effects}
{\em Conditional effects} allow planning actions to have different semantics according to the value of the current state. This feature is useful for compactly defining our method for {\em model recognition as planning}.

@@

\subsection{The observation model}
Given a classical planning problem $P=\tup{F,A,I,G}$ and a plan $\pi$ that solves $P$; the observation of the execution of $\pi$ on $P$ is $\mathcal{O}(\pi,I)=\tup{a_1^o,s_1^o \ldots , a_l^o, s_m^o}$, an interleaved combination of {\small $1\leq l\leq |\pi|$} observed actions and {\small $1\leq m\leq |\pi|$} observed states such that:
\begin{itemize}
\item Observed actions are {\em consistent} with $\pi$~\cite{ramirez2009plan}. This means that $\tup{a_1^o, \ldots, a_l^o}$ is a sub-sequence of the solution plan $\pi$.
\item Observed states are a sub-sequence  of partial states {\em consistent} with the sequence of states $\tup{s_0, s_1, \ldots, s_n}$ traversed by $\pi$. 
\end{itemize}

On the one hand, the initial state $I$ is fully observed while the observed states in $\mathcal{O}(\pi,I)$ may be partial, i.e. the value of certain fluents in the intermediate states may be omitted (~$|s_i^o|\leq |F|$ for every $1\leq i\leq m$). On the other hand, the sequence of observed states $\tup{s_1^o, \ldots, s_m^o}$ in $\mathcal{O}(\pi,I)$ is the same sequence of states traversed by $\pi$ but certain states may also be omitted. Formally, $0\leq|s_i^o|$ for every $1\leq i\leq m$. This means that the transitions between two consecutive observed states in $\mathcal{O}(\pi,I)$ may require the execution of more than a single action ($\theta(s_i^o,\tup{a_1,\ldots,a_k})=s_{i+1}^o$, where ${\small k\geq 1}$ is unknown and unbound). Therefore we can conclude that having $\mathcal{O}(\pi,I)$ does not implies knowing the actual length of $\pi$.

\begin{definition}[$\Phi$-observation]
Given a subset of fluents $\Phi\subseteq F$ we say that $\mathcal{O}(\pi,I)$ is a $\Phi$-observation of the execution of $\pi$ on $P$ iff, for every ${\small 1\leq i\leq m}$, each observed state $s_i^o$ only contains fluents in $\Phi$.
\end{definition}

Figure~\ref{fig:grid-example} illustrates the six-state $\Phi$-observation \{{\tt\footnotesize<(xcoord 2)(ycoord 1)>, <(xcoord 3)(ycoord 1)>, <(xcoord 4)(ycoord 1)>, <(xcoord 5)(ycoord 1)>, <(xcoord 5)(ycoord 2)>, <(xcoord 4)(ycoord 2)>}\}, where $\Phi$ only contains fluents of the kind {\tt\small (xcoord ?v)} and {\tt\small (ycoord ?v)}. This means that, for each observed state, only the value of fluents {\tt\small (xcoord ?v)} and {\tt\small (ycoord ?v)} is known while the value of the remaining fluents, namely {\tt\small (next ?v1 ?v2)}, {\tt\small (q0)} and {\tt\small (q1)}, is unknown.

\section{Model Recognition}
\label{sec:recognition}
The {\em model recognition} task is a tuple $\tup{P,M,\mathcal{O}}$ where:
\begin{itemize}
\item $P=\tup{F,A[\cdot],I,G}$ is a {\em classical planning problem} s.t. the semantics of actions $a\in A[\cdot]$ is unknown, i.e. the corresponding functions $\rho(s,a)$ and/or $\theta(s,a)$ are undefined.
\item $M=\{\mathcal{M}_1,\ldots,\mathcal{M}_m\}$ is a finite non-empty {\em set of models} for the actions in $A$ s.t., each model in $\mathcal{M}\in M$, defines different function pairs $\tup{\rho,\theta}$ over the state variables $F$.
\item $\mathcal{O}(\pi,I)$ is an {\em observation} of the execution of an unknown solution plan $\pi$ for the planning problem $P$.
\end{itemize}

{\em Model recognition} can be understood as a {\em classification task} where each class is represented with a different planning model $\mathcal{M}\in M$ and the observed plan execution $\mathcal{O}(\pi,I)$ is the single example to classify. Further, the planning model associated to each class acts as the corresponding {\em class prototype} that summarizes all the plan executions that could be synthesized with that model (i.e. all the examples that belong to that class).

In this work we follow the {\em naive Bayes classifier} to assign a model $\mathcal{M}\in M$ to the given observation $\mathcal{O}(\pi,I)$ with respect to the expression:
\begin{align}
argmax_{\mathcal{M}\in M} P(\mathcal{O}|\mathcal{M})\times P(\mathcal{M}).
\end{align}
The hypotheses in {\em model recognition} are then about the possible action models $\mathcal{M}\in M$, while the observation $\mathcal{O}(\pi,I)$ of a plan execution represents the input observation. The {\em solution} to the {\em model recognition} task is the model $\mathcal{M}\in M$ (or subset of models in $M$) that maximizes the previous expression.

The $P(\mathcal{M})$ probability expresses whether one model is known to be a priori more likely than the others. If this probability is not given as input it is reasonable to assume that {\em a priori} all models are equiprobable. The challenge in our formulation for {\em model recognition} is the definition of the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood} that expresses the probability of observing $\mathcal{O}(\pi,I)$ when $\mathcal{M}$ is the planning model.

\begin{figure}
  \begin{scriptsize}
  \begin{center}
	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,semithick]
	  \node[state] (A)              {$\mathcal{M}$};
	  \node[state] (B) [right of=A] {$\mathcal{M}'$};
	  \node[state] (C) [right of=B] {$\pi$};
	  \node[state] (D) [right of=C] {$\mathcal{O}$};          
	  \path
(A) edge  node {} (B)
(B) edge  node {} (C)
(C) edge  node {} (D)
;
	\end{tikzpicture}
  \end{center}
  \end{scriptsize}  
 \caption{\small {\em Bayesian network} abstracting that model $\mathcal{M}'$ results from editing $\mathcal{M}$ and that can produce a plan $\pi$ consistent with $\mathcal{O}(\pi,I)$.}
\label{fig:net}
\end{figure}

Our approach to formulate the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood} is to assess how far is $\mathcal{M}$ from a model $\mathcal{M'}$ that can produce a plan $\pi$ that solves $P$ and s.t. $\mathcal{O}(\pi,I)$ is a {\em consistent} observation of the execution of $\pi$ on $P$. Figure~\ref{fig:net} shows the four-variable {\em Bayesian network} abstracting this process. Regarding this network we have that:
\begin{align}
 P(\mathcal{O}|\mathcal{M})=\sum_{\mathcal{M}'}\sum_{\pi} P(\mathcal{M'}|\mathcal{M})P(\pi|\mathcal{M'})P(\mathcal{O}|\pi),
\end{align}
where $\mathcal{M}'$ ranges over all the models that can be generated editing $\mathcal{M}$ and $\pi$ ranges over all the plans that can be synthesized with a model $\mathcal{M}'$

\subsection{Approximating the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood}}
Following the previous equation (2) the exact computation of $P(\mathcal{O}|\mathcal{M})$ is intractable because, for most planning problems, the set of {\em valid} plans consistent with an observation can easily be very large (infinite in the case of planning problems without dead-ends). Instead, in this work we propose to estimate $P(\mathcal{O}|\mathcal{M})$ making the following assumptions:
\begin{enumerate}
\item The sum in the previous equation is dominated by the {\em closest compliant set}, i.e. the models closest to $\mathcal{M}$ that can produce a solution plan consistent with $\mathcal{O}(\pi,I)$. The rationale behind this assumption is that the further the compliant model, the lower the $P(\mathcal{O}|\mathcal{M})$ likelihood. 
\end{enumerate}

The {\em full observability of the executed plan} is a too strong assumption but, as we explain next, it allows us to build reasonable estimates of the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood} for the more general case where the observability of the executed plan is partial.

Note that in this baseline scenario where there is {\em full observability of the executed plan}:
\begin{itemize}
\item There is only a single possible plan {\em consistent} with the given observation, besides $P(\mathcal{O}|\pi)=1$.
\item If we consider assumption 1, the {\em closest compliant set} $\mathcal{M}^*$ becomes a singleton for this particular scenario, so that probabilities corresponding to different compliant models are not added up. Further, the $P(\pi|\mathcal{M^*})$ probability equals also to $1$ since there is only one possible plan consistent with the observation and by definition, the {\em closest compliant} model produces that plan.
\end{itemize}
As a consequence, when there is {\em full obsevability of the executed plan} and under the previous assumption, we have that expression (1) simplifies to
\begin{align}
argmax_{\mathcal{M}\in M} P(\mathcal{M^*}|\mathcal{M}).
\end{align}

Next we show how to approximate the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood} for the more general case where there is partial observability of the executed plan. First we start with the simpler scenario where the length of the observed plan is bound and add these two new assumptions.
\begin{enumerate}\addtocounter{enumi}{1}
\item The {\em closest compliant set} is a singleton, so that probabilities corresponding to different compliant models are not added up.
\item All solutions of a given {\em plan length} are equiprobable, with shorter plans more likely than longer plans.  
\end{enumerate}

In such an scenario $P(\mathcal{O}|\pi)$ equals to a constant $\alpha$ that indicates the number of plans of bounded length that are consistent with $\mathcal{O}$. Likewise, $P(\pi|\mathcal{M^*})$ is also one given the definition of the {\em closest compliant set}. As a consequence, when there is {\em partial obsevability of the executed plan} (but the plan length is bound), and under the previous three assumptions, we have that expression (1) simplifies to
\begin{align}
argmax_{\mathcal{M}\in M} P(\mathcal{M^*}|\mathcal{M})\times \alpha.
\end{align}
where $\alpha$ is a normalizing constant that has the same value for every $\mathcal{M}\in M$, so it can be ignored.

Finally, we can say that the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood} can be estimated for the general case of partial observability of the executed plan (where the length of the plan is not bound) by adding a fourth assumption:
\begin{enumerate}\addtocounter{enumi}{3}
\item The observed plan $\pi$ that solves $P$ is an optimal plan since we are assuming that the observed agent is acting rationally~\cite{ramirez2012plan}.
\end{enumerate}
In this case the length of the plan becomes bound by the optimal plan length and hence, we can again use expression (4) for the  likelihood estimation.

\section{Recognition of \strips\ models}
\label{sec:asPlanning}
Here we analyze the particular instantiation of the {\em model recognition} task where the semantics of actions (i.e. $\rho$ and $\theta$ functions) are specified with \strips\ action schemas. We start formalizing the \strips\ schema and define the full space of possible \strips\ schema. Eventually, we introduce an {\em edit distance} for \strips\ schema to estimate the $P(\mathcal{O}|\mathcal{M})$ likelihoods for classical planning models.

@@

A \strips\ action schema $\xi$ is defined by a list of {\em parameters} $pars(\xi)$, and three list of predicates (namely $pre(\xi)$, $del(\xi)$ and $add(\xi)$) that shape the kind of fluents that can appear in the {\em preconditions}, {\em negative effects} and {\em positive effects} of the actions induced from that schema.

We say that two \strips\ schemes $\xi$ and $\xi'$ are {\em comparable} iff $pars(\xi)=pars(\xi')$, both share the same list of parameters. For instance, we claim that the six action schema of Figure~\ref{fig:model-example} are {\em comparable} while, for example, the {\small\tt stack(?v1,?v2)} and {\small\tt pickup(?v1)} schemes from the four opertator {\em blocksworld}~\cite{slaney2001blocks} are not. Last but not least, we say that two \strips\ models $\mathcal{M}$ and $\mathcal{M}'$ are {\em comparable} iff there exists a bijective function $\mathcal{M} \mapsto \mathcal{M}^*$ that maps every action schema $\xi\in\mathcal{M}$ to a comparable schema $\xi'\in\mathcal{M'}$ and vice versa.

@@

Despite any element from ${\mathcal I}_{\Psi,\xi}$ can {\em a priori} appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists of a schema $\xi$. The space of possible \strips\ schema is bound further by ${\mathcal C}$, a set of constraints of the following kinds: 
\begin{itemize}
\item {\em Syntactic constraints}. \strips\ constraints require negative effects appearing as preconditions, negative effects cannot be positive effects at the same time and also, positive effects cannot appear as preconditions. Formally, $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$). Considering exclusively these syntactic constraints, the size of the space of possible \strips\ schema is given by the expression, $2^{2\times|{\mathcal I}_{\Psi,\xi}|}$. For the navigation model of Figure~\ref{fig:model-example}, $2^{2\times 10}=1,048,576$ for every action schema.
\item {\em Domain-specific constraints}. One can also introduce domain-specific knowledge to more precisely bound the space of possible \strips\ schema for a particular domain. For instance, in a {\em robot navigation} model, like the one in Figure~\ref{fig:model-example}, {\small\tt q0()} and {\small\tt q1()} are exclusive so they cannot hold a the same time in a $pre(\xi)$/$del(\xi)$/$add(\xi)$ list. Further, {\small\tt next($v_1$,$v_1$)} and {\small\tt next($v_2$,$v_2$)} will not appear at any of these lists because the {\tt\small next} predicate is coding the {\em succesor} function for {\em natural numbers}. In this case, introducing these domain-specific constraints reduces the size of the space of possible schema to $2^{2\times 7}=16,384$ for every action schema.
\end{itemize}

>>>>

\maketitle
\begin{abstract} 
Given a partially observed plan execution, and a set of possible planning models (models that share the same state variables but that update these variables according to different action models); {\em model recognition} is the task of identifying which model in the set produced/explains the given observation. The paper formalizes the {\em model recognition} task and introduces a novel method to estimate the probability of a \strips\ model to produce an observation of a plan execution. This method, that we called {\em model recognition as planning}, is built on top of off-the-shelf classical planning algorithms and is robust to missing intermediate states and actions in the observation of the plan execution. The effectiveness of {\em model recognition as planning} is shown in a set of \strips\ models encoding different kinds of {\em automata}. We show that {\em model recognition as planning} succeeds to identify the executed automata despite some state variables (e.g. the internal machine state) or the actual applied transitions are unobserved.
\end{abstract}

@@

This paper formalizes the {\em model recognition} task where the object to recognize is not a goal but the {\em planning model} that determines the behavior of the observed agent. Given a partially observed plan execution, and a set of possible planning models (models that share the same state variables but that update these variables with different action models); {\em model recognition} is the task of identifying the model in the set with the highest probability of producing/explaining the given observation.

\begin{figure}
  \begin{scriptsize}    
  \begin{center}
  \begin{tikzpicture}[scale=.4]
          \begin{scope}
            \draw (0, 0) grid (5, 5);
            \node[anchor=center] at (3.5, 1.5) {R};
            \node[anchor=center] at (0.5, 0.5) {};
            \draw[thick,style=dotted] (0.5,0.5) -- (4.5,0.5);
            \draw[thick,style=dotted] (4.5,0.5) -- (4.5,1.5);
            \draw[thick,style=dotted,->] (4.5,1.5) -- (3.7,1.5);                        
          \end{scope}
        \end{tikzpicture}
\hspace*{1cm}        
	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,semithick]
	  \node[state] (A)              {$q_0$};
	  \node[state] (B) [right of=A] {$q_1$};          
	  \path
(A) edge [bend left, align=center] node {inc-y,\\dec-y} (B)
(B) edge [bend left, align=center] node {inc-y,\\dec-y} (A);
	\end{tikzpicture}
  \end{center}
  \end{scriptsize}  
 \caption{\small ({\em Left)} Robot navigating a $5\times 5$ grid. {\em (Right)} Automata controlling that the robot only increments its x-coordinate at {\em even} rows (when $q_0$ holds, given that actions {\tt inc-y} and {\tt dec-y} update the robot y-coordinate and the automata state).}
\label{fig:grid-example}
\end{figure}

To better illustrate {\em model recognition}, imagine a robot in a $n\times n$ grid whose navigation is determined by the \strips\ model of Figure~\ref{fig:model-example}. According to this model the robot could increment its {\em x-coordinate} when {\tt\small $q0$} holds (i.e. at {\em even rows} if {\tt\small $q0$} holds initially) and decrement it when {\tt\small $q1$} holds (at {\em odd rows} if {\tt\small $q0$} holds initially). Apart from this particular navigation model, different action models could be defined within the same state variables (e.g. altering the way {\tt\small $q0$} and {\tt\small $q1$} are required and updated) and these models can determine different kinds of robot navigation. Given an observation of a plan execution (like the one illustrated at Figure~\ref{fig:grid-example}) {\em model recognition} would aim here to identify which navigation model produced/explains that observation, despite key information is unobserved (e.g. the particular applied actions or the value of state variables {\small\tt $q0$} and {\small\tt $q1$}).

{\em Model recognition} is of interest because once the planning model is recognized, then the model-based machinery for automated planning becomes applicable~\cite{ghallab2004automated}. In addition, it enables identifying different kinds of automata by observing their execution. It is well-known that diverse automata representations, like {\em finite state controllers}, {\em push-down automata}, {\em {\sc GOLOG} programs} or {\em reactive policies}, can be encoded as classical planning models~\cite{baier2007exploiting,Geffner:FSM:AAAI10,ivankovic2015optimal,segovia2017generating}.

The paper also introduces {\em model recognition as planning}; a novel method to estimate the probability of a given \strips\ model to produce an observed plan execution. Our method is built on top of off-the-shelf classical planning algorithms and is robust to missing intermediate states and actions in the observed plan execution. We evaluate the effectiveness of {\em model recognition as planning} with sets of \strips\ models that represent different {\em automata}. All the {\em automata} in a set are defined within the same state variables but different {\em transition functions}. We show that {\em model recognition as planning} succeeds to identify the executed {\em automata} despite some state variables (e.g. the internal machine state) or the actual applied transitions are unobserved.

@@

A {\em classical planning frame} is a $\tup{F,A}$ pair, where $F$ is a set of {\em fluents} and $A$ is a set of {\em actions}. The semantics of actions $a\in A$ is specified with two functions: $\rho(s,a)$ that denotes whether the action $a$ is {\em applicable} in a state $s$ and $\theta(s,a)$ that denotes the {\em successor state} that results of applying the action $a$ in a state $s$. In this work we compactly represent the $\rho$ and $\theta$ functions following the \strips\ model. With this regard, an action $a\in A$ is defined by:
\begin{itemize}
\item $\pre(a)\in\mathcal{L}(F)$, the {\em preconditions} of $a$, is the set of literals that must hold for the action $a\in A$ to be applicable.
\item $\eff^+(a)\in\mathcal{L}(F)$, the {\em positive effects} of $a$, is the set of literals that are true after the application of action $a\in A$.
\item $\eff^-(a)\in\mathcal{L}(F)$, the {\em negative effects} of $a$, is the set of literals that are false after the application of the action.
\end{itemize}
We say that an action $a\in A$ is {\em applicable} in a state $s$ iff $\pre(a)\subseteq s$. The result of applying $a$ in $s$ is the {\em successor state} denoted by $\theta(s,a)=\{s\setminus\eff^-(a))\cup\eff^+(a)\}$.

@@

\subsection{Conditional effects}
{\em Conditional effects} allow classical planning actions to have different semantics according to the value of the current state. This feature is useful for compactly defining our method for {\em model recognition as planning}.

@@

\subsection{The observation model}
Given a classical planning problem $P=\tup{F,A,I,G}$ and a plan $\pi$ that solves $P$; the observation of the execution of $\pi$ on $P$ is $\mathcal{O}(\pi,P)=\tup{a_1^o,s_1^o \ldots , a_l^o, s_m^o}$, an interleaved combination of {\small $1\leq l\leq |\pi|$} observed actions and {\small $1\leq m\leq |\pi|$} observed states such that:
\begin{itemize}
\item Observed actions are {\em consistent} with $\pi$~\cite{ramirez2009plan}. This means that $\tup{a_1^o, \ldots, a_l^o}$ is a sub-sequence of the solution plan $\pi$.
\item Observed states are a sequence of partial states that is {\em consistent} with the sequence of states $\tup{s_0, s_1, \ldots, s_n}$ traversed by the execution of $\pi$ on $P$. 
\end{itemize}

On the one hand, the initial state $I$ is fully observed while the observed states in $\mathcal{O}(\pi,P)$ may be partial, i.e. the value of certain fluents in the intermediate states may be omitted (~$|s_i^o|\leq |F|$ for every $1\leq i\leq m$). On the other hand, the sequence of observed states $\tup{s_1^o, \ldots, s_m^o}$ in $\mathcal{O}(\pi,P)$ corresponds to the same sequence of states traversed by $\pi$ but certain states may also be omitted ($0\leq|s_i^o|$ for every $1\leq i\leq m$). This means that the transitions between two consecutive observed states in $\mathcal{O}(\pi,P)$ may require the execution of more than a single action ($\theta(s_i^o,\tup{a_1,\ldots,a_k})=s_{i+1}^o$, where ${\small k\geq 1}$ is unknown and unbound). Therefore we can conclude that having $\mathcal{O}(\pi,P)$ does not imply knowing the actual length of $\pi$.

\begin{definition}[$\Phi$-observation]
Given a subset of fluents $\Phi\subseteq F$ we say that $\mathcal{O}(\pi,P)$ is a $\Phi$-observation of the execution of $\pi$ on $P$ iff, for every ${\small 1\leq i\leq m}$, each observed state $s_i^o$ only contains fluents in $\Phi$.
\end{definition}

Figure~\ref{fig:grid-example} illustrates the six-state $\Phi$-observation \{{\tt\scriptsize<(xcoord 2)(ycoord 1)>, <(xcoord 3)(ycoord 1)>, <(xcoord 4)(ycoord 1)>, <(xcoord 5)(ycoord 1)>, <(xcoord 5)(ycoord 2)>, <(xcoord 4)(ycoord 2)>}\}. In this case $\Phi$ only contains fluents of the kind {\tt\small (xcoord ?v)} and {\tt\small (ycoord ?v)} so that the value of the remaining fluents, namely {\tt\small (next ?v1 ?v2)}, {\tt\small (q0)} and {\tt\small (q1)}, is unknown.

\section{Model Recognition}
\label{sec:recognition}
The {\em model recognition} task is a tuple $\tup{P,M,\mathcal{O}}$ where:
\begin{itemize}
\item $P=\tup{F,A[\cdot],I,G}$ is a {\em classical planning problem} s.t. the semantics of actions $a\in A[\cdot]$ is unknown (i.e. the corresponding $\rho$ and/or $\theta$ functions are undefined).
\item $M=\{\mathcal{M}_1,\ldots,\mathcal{M}_m\}$ is a finite non-empty {\em set of models} for the actions in $A[\cdot]$ s.t., each model in $\mathcal{M}\in M$, defines different function pairs $\tup{\rho,\theta}$ over the state variables $F$.
\item $\mathcal{O}(\pi,P)$ is an {\em observation} of the execution of an unknown solution plan $\pi$ for the planning problem $P$.
\end{itemize}

{\em Model recognition} can be understood as a {\em classification task} where each class is represented with a different planning model $\mathcal{M}\in M$ and the observed plan execution $\mathcal{O}(\pi,P)$, is the single example to classify. The planning model associated to each class acts as the corresponding {\em class prototype} and summarizes any observation of a plan execution that could be synthesized with that model (i.e. all the examples that belong to that class).

We follow the {\em naive Bayes classifier} to assign a model $\mathcal{M}\in M$ to the given observation $\mathcal{O}(\pi,P)$ with respect to the expression:
\begin{align}
argmax_{\mathcal{M}\in M} P(\mathcal{O}|\mathcal{M})\times P(\mathcal{M}).
\end{align}
The hypotheses in {\em model recognition} are then about the possible action models $\mathcal{M}\in M$, while the input observation represents the observation $\mathcal{O}(\pi,P)$ of the execution of a plan $\pi$ that solves $P$. The {\em solution} to the {\em model recognition} task is the subset of models in $M$ that maximizes the previous expression (1).

\subsection{Formulating the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood}}
The $P(\mathcal{M})$ probability expresses whether one model is known to be a priori more likely than the others. If this probability is not given as input it is reasonable to assume that, {\em a priori}, all models are equiprobable. The challenge in our formulation for {\em model recognition} is the definition of the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood} that expresses the probability of observing $\mathcal{O}(\pi,P)$ when $\mathcal{M}$ is the planning model.

\begin{figure}
  \begin{scriptsize}
  \begin{center}
	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,semithick]
	  \node[state] (A)              {$\mathcal{M}$};
	  \node[state] (B) [right of=A] {$\mathcal{M}'$};
	  \node[state] (C) [right of=B] {$\pi$};
	  \node[state] (D) [right of=C] {$\mathcal{O}$};          
	  \path
(A) edge  node {} (B)
(B) edge  node {} (C)
(C) edge  node {} (D)
;
	\end{tikzpicture}
  \end{center}
  \end{scriptsize}  
 \caption{\small {\em Bayesian network} abstracting that model $\mathcal{M}$ can be {\em transformed} into a model $\mathcal{M}'$ that is able to produce a plan $\pi$ consistent with observation $\mathcal{O}(\pi,P)$.}
\label{fig:net}
\end{figure}

Our approach to formulate the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood} is to assess the cost of {\em transforming} $\mathcal{M}$ into a model $\mathcal{M'}$ that can produce a plan $\pi$ that solves $P$ and s.t. $\mathcal{O}(\pi,P)$ is {\em consistent} with $\pi$. Figure~\ref{fig:net} shows the four-variable {\em Bayesian network} abstracting this process. Regarding this network we have the following formulation:
\begin{align}
 P(\mathcal{O}|\mathcal{M})=\sum_{\mathcal{M}'}\sum_{\pi} P(\mathcal{M'}|\mathcal{M})P(\pi|\mathcal{M'})P(\mathcal{O}|\pi),
\end{align}
where $\pi$ ranges over all the plans that can be synthesized with a model $\mathcal{M}'$ and $\mathcal{M}'$ ranges over all the models that can be generated {\em transforming} $\mathcal{M}$.

Edit distances are similarity metrics, traditionally computed over {\em strings} or {\em graphs}, and that has been proved successful for {\em pattern recognition}~\cite{masek1980faster,bunke1997relation}. We assess the cost of {\em transforming} a classical planning model $\mathcal{M}$ into a model $\mathcal{M'}$ formalizing and computing {\em edit distances} that, in this work, are referred to \strips\ planning models.

\section{Recognition of \strips\ models}
\label{sec:asPlanning}
Here we analyze the particular instantiation of the {\em model recognition} task where the semantics of actions (i.e. $\rho$ and $\theta$ functions) are specified with \strips\ action schema. We start formalizing the \strips\ schema and define the full space of possible \strips\ schema. Eventually, we introduce an {\em edit distance} for \strips\ schema to estimate the $P(\mathcal{O}|\mathcal{M})$ likelihoods for classical planning models.

@@

{\em A \strips\ action schema} $\xi$ is defined by a list of {\em parameters} $pars(\xi)$, and three list of predicates (namely $pre(\xi)$, $del(\xi)$ and $add(\xi)$) that shape the kind of fluents that can appear in the {\em preconditions}, {\em negative effects} and {\em positive effects} of the actions induced from that schema.

We say that two \strips\ schemes $\xi$ and $\xi'$ are {\em comparable} iff $pars(\xi)=pars(\xi')$, both share the same list of parameters. For instance, we claim that the six action schema of Figure~\ref{fig:model-example} are {\em comparable} while, for example, the {\small\tt stack(?v1,?v2)} and {\small\tt pickup(?v1)} schemes from the four operator {\em blocksworld}~\cite{slaney2001blocks} are not. Last but not least, we say that two \strips\ models $\mathcal{M}$ and $\mathcal{M}'$ are {\em comparable} iff there exists a bijective function $\mathcal{M} \mapsto \mathcal{M}^*$ that maps every action schema $\xi\in\mathcal{M}$ to a comparable schema $\xi'\in\mathcal{M'}$ and vice versa.

@@

Despite any element from ${\mathcal I}_{\Psi,\xi}$ can {\em a priori} appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of a schema $\xi$, the space of possible \strips\ schema is constrained further by ${\mathcal C}$ which includes: 
\begin{itemize}
\item {\em Syntactic constraints}. \strips\ constraints require negative effects appearing as preconditions, negative effects cannot be positive effects at the same time and also, positive effects cannot appear as preconditions. Formally, $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$). Considering exclusively these syntactic constraints, the size of the space of possible \strips\ schema is given by $2^{2\times|{\mathcal I}_{\Psi,\xi}|}$. For every action schema in the navigation model of Figure~\ref{fig:model-example} then $2^{2\times 10}=1,048,576$.
\item {\em Domain-specific constraints}. One can also introduce domain-specific knowledge to more precisely constrain the space of possible schema for a particular domain. For instance in a {\em robot navigation} model like the one in Figure~\ref{fig:model-example}, {\small\tt q0()} and {\small\tt q1()} are exclusive so they cannot hold at the same time in a $pre(\xi)$/$del(\xi)$/$add(\xi)$ list. Further, {\small\tt next($v_1$,$v_1$)} and {\small\tt next($v_2$,$v_2$)} will not appear at any of these lists because the {\tt\small next} predicate is coding the {\em successor} function for {\em natural numbers}. These domain-specific constraints reduces the size of the space of possible schema to $2^{2\times 7}=16,384$ for every action schema.
\end{itemize}
